---
title: "ANOVA for Journalists: Analyzing College Admissions Test Scores"
date: 11-3-2025
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this exercise, we'll explore a common journalism question: *Are there significant differences in standardized test scores across different types of colleges?*

We'll use ANOVA (Analysis of Variance) to determine if the observed differences in SAT scores across three college types (Public, Private, and Community College) represent real patterns or just random variation.

## Task 1: Load Libraries and Create Sample Data

```{r}
library(tidyverse)
```

```{r create-data}
# Set seed for reproducibility
set.seed(123)

# Create simulated SAT scores for admitted students across three college types
# Scale is 400-1600 for total SAT
# rnorm generates random numbers, in this case, 100 numbers
public_college <- rnorm(100, mean = 1350, sd = 120)
private_college <- rnorm(100, mean = 1450, sd = 100)
community_college <- rnorm(100, mean = 1050, sd = 150)

# Combine into a single dataframe
admissions_data <- data.frame(
  college_type = factor(rep(c("Public", "Private", "Community"), each = 100)),
  sat_score = c(public_college, private_college, community_college)
)

# Ensure SAT scores are within realistic range (400-1600)
admissions_data$sat_score <- pmin(pmax(admissions_data$sat_score, 400), 1600)

# Quick look at the data
head(admissions_data)
```

## Task 2: Calculate Basic Statistics

```{r summary-stats}
# Summary statistics by college type
admission_summary <- admissions_data |>
  group_by(college_type) |>
  summarize(
    mean_score = mean(sat_score),
    median_score = median(sat_score),
    sd_score = sd(sat_score),
    min_score = min(sat_score),
    max_score = max(sat_score)
  )

print(admission_summary)
```

Context: Fall 2024 midpoint SAT for University of Maryland was 1,465.


**Task 3** Describe the data

Write a couple of sentences describing the differences in SAT scores between types of colleges, using the descriptive statistics we just generated.


**The SAT scores vary depending on the type of college. The mean score at the three colleges ranked private as the highest with 1435, public as second with 1360, and community as third with 1068. The differences are further exemplified through the minimum, which shows private school as the highest with 1244, public as the second with 1072, and community as third with 786. The median and standard deviation scores show the same pattern, which is that private schools have the highest score, public schools come next, and community has the lowest scores.**


## Task 4: Visualize the Data

Let's create visualizations to see the patterns:

```{r visualize-distributions}
# Create density plot showing overlapping distributions
ggplot(admissions_data, aes(x = sat_score, fill = college_type)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "SAT Score Distributions by College Type",
    subtitle = "Are these differences statistically significant?",
    x = "SAT Score (400-1600 scale)",
    y = "Density",
    fill = "College Type"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

```{r boxplot}
# Create boxplot for another view
ggplot(admissions_data, aes(x = college_type, y = sat_score, fill = college_type)) +
  geom_boxplot() +
  labs(
    title = "SAT Scores by College Type",
    x = "College Type",
    y = "SAT Score"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

## Task 5: Perform ANOVA Test

```{r anova-test}
# Run one-way ANOVA
anova_result <- aov(sat_score ~ college_type, data = admissions_data)
summary(anova_result)
```

The F-value is huge (284.3). The F-value measures the ratio of between-group variance to within-group variance, and this high value suggests the differences between college types far outweigh the variation within each type. The p-value is very low, indicating the differences between college types are statistically significant at any reasonable significance level. The three asterisks (***) highlight that this is significant at the 0.001 level. So we can say that with confidence that there are real differences in SAT scores between the three types of colleges.

How do we know the F-value is huge?
We can calculate the critical value for an F statistic by using the qf() function in R

```{r}
# Your ANOVA has:
# df1 (between groups) = 2
# df2 (within groups/residuals) = 297
# Common alpha levels: 0.05, 0.01, 0.001

# Critical values at different significance levels:
critical_f_05 <- qf(0.95, df1 = 2, df2 = 297)
critical_f_01 <- qf(0.99, df1 = 2, df2 = 297)
critical_f_001 <- qf(0.999, df1 = 2, df2 = 297)

cat("Here are the Critical F-values:\n")
cat("alpha of α = 0.05: ", round(critical_f_05, 3), "\n")
cat("alpha of α = 0.01: ", round(critical_f_01, 3), "\n")
cat("alpha ofα = 0.001:", round(critical_f_001, 3), "\n")
cat("The F-statistic is: 284.3\n")
cat("Ratio to most extreme critical value (α=0.001) to the F statistic:", round(284.3/critical_f_001, 1), "x\n")
```


## Task 5: Compare the differences between specific pairs

```{r post-hoc}
# Tukey's HSD test to see which specific pairs of college types differ
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)
```

The Tukey multiple comparisons test shows highly significant differences in SAT scores between all college types:

* Private vs. Community: Private colleges have SAT scores that are 366.96 points higher than community colleges on average (p < 0.0001)
* Public vs. Community: Public colleges have SAT scores that are 292.55 points higher than community colleges on average (p < 0.0001)
* Public vs. Private: Public colleges have SAT scores that are 74.40 points lower than private colleges on average (p = 0.000021)

All comparisons show extremely small p-values, indicating these differences are statistically significant at any conventional significance level. The 95% confidence intervals (shown in the "lwr" and "upr" columns) do not include zero for any comparison, further confirming these differences are statistically significant.

## Task 6: Report the Findings

Based on our analysis, we can now answer:

1. Is there a significant difference in SAT scores across college types?
There is a statistically significant difference in SAT scores across college types. The comparisons in the Tukey test show this with the small p values for all three categories. It is less than 0.0001 for public v. community and private v. community, and equal to 0.000021 for public v. private. The ANOVA test produced a very small p value of 2e.16, which indicates that not all colleges have the same SAT scores. It is unlikely that these are due to chance, as shown by the tests. 

2. Which specific college types differ from each other?
The college types all differ from one another, but the public v private has the most significant difference. The variance of public v. private makes the 74 point difference produce a p value of 0.000021. There is a 95% confidence interval which confirms the differences. 

3. How large are these differences?
The differences are very large, as shown in the Tukey Test. Private v. Community has a large difference of 366.956, public v. community has a difference of 292.55, and public v. private has a difference of -74.404. The gaps show that private schools are at the top, with community colleges at the bottom.  

Answer those questions, providing evidence for your answers, and then write a headline and a lead paragraph reporting these findings using active voice and focusing on the differences across types.

Private Schools have the highest SAT score when compared to public and community schools.

Three types of schools showed statistically significant results when compared against one another for a difference in SAT scores. Private schools had the highest average score, while community schools had the lowest average score. When public school scores were compared to private school scores, the data showed that there was the most significant difference between these types, with a multiple determining values proving the significance. 



