---
title: "Diversity Index for Maryland Cities"
date: "Nov 8, 2025"
author: "Ashley Burkett"
---

**Ashley Burkett**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?



```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

**The data contains the names of each city in Maryland, along with their total population count. This population count is broken down into individual races such as White, Black, Native, and Asian, along with a category called 2+ races. The city that stands out as the most diverse to me based off of this data is Baltimore City, due to their numbers being the highest for almost each category. However, I know that their population is also the largest. In order to know what's the most diverse city based on percentages, I need to do more calculations. **


## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?


```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```

**The city with the highest race probability is Summerfield. This is because nearly the entire population is Black. The total population is 12403, and the number Black people in that city is 11565. This makes the probability that someone in Summerfield will be Black very high, since those numbers are so close.**

## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?


```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```


**The cities with the lowest Hispanic probability score also have a low race probability score. This means that areas with little diversity also have a small Hispanic population. On the opposing end, the higher Hispanic scores also show higher race probability scores.**

## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?


```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

**The cities with the highest diversity score are Langley Park (1), Wheaton (2), and Adelphi (3). They are all located very close to College Park. The three least diverse cities are Lake Shore (128), Lake Arbor (127), and Largo (126). They are all to the right of Washington D.C, and closer to the water. Each city is more east than the three most diverse cities.**

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

**Hubs of Diversity in Maryland Cities, and Lack of it, are Both Grouped Near Washington D.C.**

**A diversity index showed that the most diverse, and least diverse, cities in Maryland are clustered in the outskirts of Washington D.C. Census race and ethnicity data from 2020 was used to calculate the Diversity Index across cities in Maryland. Langley Park sits at the most diverse city, with an index of 97.18. There is just a 3% chance that two people would be of the same race or ethnicity. On the other end of the spectrum, Lake Shore dives down to an index of just 15.53. The chance that two people will not be of the same race or ethnicity is just one in seven.**

**To complete this story, I would talk to locals in each of the least and most diverse cities. It's especially important to talk to residents of Langley Park and Lake Shore, given they are on opposite ends of the diversity spectrum. I would ask about how they perceive diversity in their community, or lack thereof. I would also consider talking to community leaders or legislative officials to discuss the effects of diversity in their community, from an economic and politial standpoint.**

