---
title: "Margin or Error / Home Values After Transit Line"
date: 10-16-2025
---

Activity

Divide into groups, and one of you should have access to a calculator (Google's calculator is fine). Here's your scenario:

You've received these poll results and are assigned to write a story about them. For each scenario, decide if the headline 'Majority Supports New City Policy' is statistically justified.

Poll Scenarios:

52% support, sample size 100 = 10
53% support, sample size 800 = 3.5
55% support, sample size 400 = 5
For each of these scenarios, you'll calculate the margin of error - the plus/minus for that support percentage. The formula is:

MoE = 1/√n × 100

Where:

MoE is the margin of error (expressed as a percentage)
√n is the square root of the sample size
Calculate the margin of error.
**Submit the margin of error results here:**
MoE n=100 : +/- 10
MoE n=800 : +/- 3.5
MoE n=400 : +/- 5

(ORIGNAL PERCNET X MARGIN OF ERROR) - MARGIN OF ERROR

Answer the following questions:

--  Is the headline 'Majority Supports New City Policy' statistically justified?
--  What alternative headline would be more accurate?
**Write your answers here:**

The headline is not statistically justified. This is because for n=800, which is the biggest sample size and therefore the most accurate, the percent support falls below 50%, at 49.5. 

The alternative headline is, "The Town is Split on the New City Policy."

-------------------------------------------------------------------------
# Measuring the Impact of Transit on Housing Prices
-------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(tidyverse)
```


Local officials have claimed that a new light rail line has increased property values in a local neighborhood.

Before the transit line opened, homes in that neighborhood sold for an average of $285,000, with a standard deviation of $32,000.

Since the extension opened, your news team collected sales data from 14 recent home sales in the neighborhood:
$297,000, $305,000, $292,000, $318,000, $290,000, $311,000, $299,000, $325,000, $302,000, $288,000, $307,000, $315,000, $320,000, $294,000

As a journalist, you need to determine: **Is there statistical evidence that the transit line has actually increased home values?**

Follow along, running the code when directed. If you have questions, ask either in class or put them just below the section you have a question about.

## Task 1: Organize your data

```{r}
# Known information about home values before the transit line
prior_mean <- 285000  # average home value
prior_sd <- 32000     # standard deviation

# Home sale prices after the transit line opened (14 homes)
new_prices <- c(297000, 305000, 292000, 318000, 290000, 311000, 299000, 
                325000, 302000, 288000, 307000, 315000, 320000, 294000)

# Create a journalist-friendly dataset
home_data <- tibble(
  sale_id = 1:14,
  sale_price = new_prices
)

# View the data
home_data
```

## Task 2: Calculate key statistics

```{r}
# Calculate statistics on the new home price data
new_stats <- home_data |> 
  summarise(
    mean = mean(sale_price),
    sd = sd(sale_price),
    n = n(),
    min = min(sale_price),
    max = max(sale_price)
  )

new_stats
```

## Task 3: Make a chart

```{r}
# Function to format currency values
dollar_format <- function(x) {
  paste0("$", format(x/1000, big.mark=","), "K")
}

# Create a visualization of home prices with reference to previous average
ggplot(home_data, aes(x = reorder(sale_id, sale_price), y = sale_price)) +
  geom_col(fill = "seagreen", alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 3, y = prior_mean - 8000, 
           label = "Previous Average ($285K)", hjust = 0, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = dollar_format) +
  labs(
    title = "Home Sales After Transit Line",
    subtitle = "Horizontal line shows previous neighborhood average of $285,000",
    x = "Recent Home Sales",
    y = "Sale Price",
    caption = "Source: County Property Records"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

## Task 4: Perform a hypothesis test

This is where we formally test the officials' claim that the transit line has increased home values.

**Hypotheses:**
- H₀: μ ≤ $285,000 (The transit line has not increased home values)
- H₁: μ > $285,000 (The transit line has increased home values)

```{r}
# Perform a one-sample t-test
# Since we want to know if home values increased, we use a one-sided test (alternative = "greater")
t_test_result <- t.test(
  home_data$sale_price,
  mu = prior_mean,
  alternative = "greater"
)

# Display the results
t_test_result
```

## Task 5: Interpreting the results for your story

Here's a formatting trick to better read the p value without the scientific notation. I find it easier to eyeball the results and compare with the 0.05 p value parameter
```{r}
p_value <- t_test_result$p.value

format(p_value, scientific = FALSE)

```



```{r}
# Extract the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval
ci <- t.test(home_data$sale_price)$conf.int

# Format and display the confidence interval in a readable way
ci_formatted <- tibble(
  `Lower bound` = dollar_format(ci[1]),
  `Upper bound` = dollar_format(ci[2]),
  `Confidence level` = "95%"
)

# Display as a nice table
ci_formatted

# Set our significance level
alpha <- 0.05

# Compare and make a decision
if(p_value < alpha) {
  conclusion <- "The data provides statistical evidence supporting the claim that the transit line has increased home values."
} else {
  conclusion <- "The data does not provide strong statistical evidence that the transit line has increased home values."
}

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = dollar_format(prior_mean),
  `New average` = dollar_format(mean(new_prices)),
  `Increase` = dollar_format(mean(new_prices) - prior_mean),
  `Percent change` = paste0(round(((mean(new_prices) - prior_mean) / prior_mean) * 100, 1), "%"),
  `p-value` = p_value,
  p_value_formatted = format(p_value, scientific = FALSE),
  `Statistically significant?` = if_else(p_value < alpha, "Yes", "No")
)

# Display the key statistics
story_stats

print(conclusion)

```

## Task 6: What's the story?

After disucssion the results, each person writes a couple of sentences for the top of a story about local officials' claims on the impact of the transit line and what the data says. Remember that there are lots of possible reasons that could influence the price of a home. Make your sentences newsy.

Locals near the new transit line have said that home prices are increasing drastically, and newly published data proves them right. The previous sale average of $258,000 has gradully risen to $304,500 in recent home sales, according to local county records. This is a $46,500 increase in the area. 

It could be claimed that other factors could contribute to the housing rise, but the relationship between the two variables was proven significant by a statistical tes.  

