---
title: "Car Thefts Significance"
author: "Derek Willis/Rob Wells"
date: 10-27-2025
---

name: Ashley Burkett


Your editor wants to know if there have been recent changes in the number of car thefts in Prince George's County. Using monthly figures and significance testing, you'll come up with an answer and decide whether there's a story here. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# LOAD THE TIDYVERSE
library(tidyverse)
```


### Task 2: Load the Prince George's county crime data and generate car thefts totals by month

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv")  |> 
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. 

```{r}

car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2025)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```


**Q3:** What's your initial reaction to the change in car thefts? Is there a trend?

**My initial reaction to the change in car thefts is that it is very sporadic, but there is a clear downward trend. In July 2023, there was the highest amount of car thefts at 673. It then climbs down over the next year and a half, so February 2025, where it reaches a low of 248.**


### Task 4: Add the mean, standard error and calculate a margin of error based on a 95% confidence interval

Let's add some stats to that process - mean and standard deviation - and calculate the standard error and 95% confidence intervals. We only have a small number of months, so we'll calculate the margin of error to account for that. 

**INSERT CORRECT VARIABLE AT "REPLACE ME"**

```{r}

theft_summary <- car_thefts |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,  # Using t-distribution for small sample
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

# Display the confidence interval
theft_summary
```




### Task 5: Visualize the data

Let's make a better chart showing the changes and the statistics we produced in Task 4. Does this change anything about your answer from Task 3?



```{r}
# Make a better chart
car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper), 
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month), 
           y = theft_summary$mean_thefts + 10, 
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)), 
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 6: Describe the confidence interval.

- What does the 95% confidence interval show here? **The confidence interval shows that we are confident that 95% confident that the true mean is within the red shaded area. The area depicted is between about 410 and 505. **

- What figure(s) accurately reflects the number of car thefts in a typical month? **The figure that accurately reflects the number of car thefts in a typical month is the mean.This number is 460.4 car thefts per month.**

- Propose a headline about the trend in car thefts that is focused on the present and future, but informed by the data you see above. **Car Thefts in Prince George's County Steadily Decline After a Spike in 2023**



### Task 7: Is the recent trend significant?

It _looks_ like car thefts are decreasing lately. But is that change real? We'll split the monthly data into two halves, and then ask if whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05 and develop a null hypothesis and an alternative hypothesis for this scenario.

Null Hypothesis: there is no real difference between the two periods
Alternative Hypothesis: there is a real difference between the two periods

**INSERT CORRECT VARIABLES AT "REPLACE ME"**

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |> 
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods - this is a two-sided test because we have two periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

# Visualize the comparison
period_summary <- car_thefts |> 
  group_by(period) |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months- 1) * se_thefts, # look at how we calculated this above
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |> 
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half of Data Period",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 3)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Task 8: Reflection

Given the above chart, is the difference between the two periods statistically significant? Can we reject the null hypothesis? **The difference between the two periods is not statistically significant. The P value of 0.179 is higher than the alpha value 0.05, meaning that the data is not statistically significant. The graph also shows that there is a more minor difference in car thefts between the two months, than how drastic the downward curve was in the previous chart. The null hypothesis cannot be rejected.**

How would you revise your headline from Task 6, if at all? What would you tell your editor, and what would you expect her to ask you? **My new headline would read: Prince George's County Car Thefts Appear to Take a Steep Decline, but the Numbers Don't Prove It." I would tell my editor that the data is not significant and it should not be published. I would expect her to ask me how I know that the data isn't statistically significant, and I would explain the T-Test and my P-Value.**





When you are finished, add, commit and push your notebook and submit the URL in ELMS.
