---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Ashley Burkett"
date: 8-24-2025
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

*The correlation matrix shows that those with a bachelor degree or higher, and income levels, have a strong relationship to voter turnout. This is visually depicted by darker blue, slightly larger, circles on the matrix. The variable that doesn't have much correlation is unemployment. This category has small, pale, circles showing little relationship to voter turnout.*
*The scatterplot reinforces these conclusions. A bachelors degree or higher and high income levels, both have positive correlation with voter turnout. The line rises to the right and has dots clustered near it. The unemployment category has a negative correlation with voter turnout, shown by a line dipping to the right.*
*When looking at the scatterplots, the categories with the strongest linear relationships are bachelors degree, income, and median age. Population and unemployment have negative correlations and don't show any meaningful relationship to voter turnout.*



```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

*The multiple variable model predicts voter turnout moderately well, but there are still some concerning factors about the result of the data. The median and residuals are all very close to 0, showing that the model is quite balanced and not experiencing any significant weighting. The multiple R-squared value is also 0.85, meaning that 85% of the variability is explained by the model. The p-value of 2.999e-06 is much smaller than 0.05, and shows high significance of the model as well. The categories with asterisks showing correlation are age, race (white), and most notably income. However, there are other categories that aren't noted as significant. Population, bachelors or higher, and unemployment all have no correlation to voter turnout. Their values are all over 0.05, and not significant*
*I would say this is a useful model due to the results from the p-value, residuals, median, and R-squared value. All of these results point towards significance and correlation with certain values on the model*

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

*I chose to remove each one of the non-significant variables for one of each model. In the first model I removed population, in the second model I removed bachelors degree, and in the third model I removed unemployed. Model #3 performed slightly better than the other ones, which is where I removed bachelors degree or higher. In the previous model, bachelors degree had a p-value of  0.83677, showing its insignificance. The new model has a higher R-squared value than the original, and tops the other two models that I created. Model 3 had an R-squared value of 0.8127754. Just behind is model #2 with 0.8123836, where population was removed. Model #4 had the lowest R-squared value, showing that it has the least correlation, with 0.7890148. To concluse, model #3 where I removed the category bachelors degree or higher performed significantly better than the other models.*

```{r}
# Model 2
model2 <- lm(turnout ~ income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ population + income + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ population + income + pct_bachelors_higher + 
                  median_age + pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

*Data analysis for voter turnout across Maryland counties shows that certain demographic factors affect how likely residents of each county are to vote. Several demographics were compared across each other; population of the county, income, bachelors degree or higher, median age, and percent of the population that is white. Each of these categories were measured in comparison to voter turnout. The data analysis for voter turnout found that counties with higher income and a higher percent of white residents correlate with voter turnout. This means that wealthier counties with a higher population of white community members are more likely to have a higher voter turnout.*