---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv") |> clean_names()
```
### Task 3: Describe the Data

#You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

This dataframe shows all crime in Prince George's County in a given year, by month and by incident. The chart includes types of property crimes such as theft, theft from auto and auto stolen. The chart also includes many types of violent crime such as assault, robbery and homicide. 

Categories that are more important than others would include those with high instances per month. Accident, auto stolen and theft all have very high occurrences per month, and may indicate a trend for crime in the area. This could lead to a story. 

The category, robbery residential, makes it hard to summarize the data because there is a missing data point that's filled with N/A. The chart also includes categories such as "theft" and "theft from auto." It makes it difficult to know if some incidents are double-crossed with one another across two categories. 

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> mutate(total_incidents = rowSums(across(2:13))) |>
mutate(average_incidents = total_incidents/12)
```

### Task 5: Describe the Averages

#Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The average number of incidents,for each category per month are shown in "average incidents." This column added together all of the previous incidents and divided them by 12, which gave an average. Things like auto stolen, accident and theft all have high incidence rates that correlate with the higher number of occurrences depicted in each month. The same goes for the lower incidence crimes such as homicide, breaking and entering other and sex offense. 

Accident has the highest incidence rate with 481.33 per month, and homicide has the lowest at 5.833. Theft, theft from auto and auto stolen were all very high. The crime trend in Pirnce George's county points towards a high incidence rate for motor vehicles in all situations. Violent crimes tend to stay lower. 

Robbery residential does not have a complete average column because there are missing datapoints. I wonder what their average would be if the N/A was replaced with 0, so that there could still be an accurate calculation.

### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |>
mutate(
  june_change = ((jun - average_incidents) / average_incidents) * 100,
    july_change = ((jul - average_incidents) / average_incidents) * 100,
    august_change = ((aug - average_incidents) / average_incidents) * 100
)
```

### Task 7: Describe the Summer Crime Averages

#Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

The percent change in the average number of incidents per month over the summer shows that there's no major trend over the summer months. The only crimes that experience a constant increase over the summer are theft from auto, homicide and auto stolen. The rest of the categories experience some months where the incidence rate rises, and others where it falls. 

I don't see any patterns in this data. This tells me that the theory that crime is more liekly to occur in the summer months is false. 

### Task 8: What's the Story?

#Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Vehicle related crimes are the most frequent in Prince George's County, and far exceed the occurrence of violent crimes in the area.

Crime data shows that incidents involving automobiles are among the highest in the county. Vehicles being stolen and broken into have two of the highest incidence rates, with 4,790 vehicles being stolen, and 4,703 vehicles being broken into. These far differ from homicide, with just 70 instances recorded. Outside of property crimes related to vehicles, other car related occurrences were also high. Accidents had the most total reports with 5,776 incidents. These automobile related instances show law enforcement and policymakers that there needs to be steps taken to protect vehicles and their drivers in Prince George's County. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
